---
title: "code_inference"
author: "Roy Gabriel"
output: html_document
date: "2024-06-11"
---

# Load Necessary Libraries and Data

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


library(tidyverse)
library(caret)
library(glmnet)

orbit <- read.csv("comet_data_cleaned.csv")
print(head(orbit))
```

# Variable Selection
## Forward Stepwise Regression
```{r}
#RQ1: Which orbital elements best predict the minimum distance between the comet and Earth (MOID)? Are all input parameters significant? Is the MOID impacted more by the orbit eccentricity (e) or its critical angles (i, w, Node)?

#Sub-setting data for clarity in linear regression modeling
#Deleted columns include Object, Ref, and Object_name since they will only lead to NANs if included in the model.

sub.orbit <- orbit[,2:11]
simple_model <- lm(MOID..AU. ~ 1, data=sub.orbit)
full_model <- lm(MOID..AU. ~., data = sub.orbit)

rq1_model_forward <- step(simple_model, scope=list(lower=simple_model, upper=full_model), direction="forward")
```

```{r}
summary(rq1_model_forward)
```

Based on the output of the forward stepwise regression, the variables selected were `Q..AU`, `P..yr`, `q..AU`, `e`, `Node..deg`, and `w..deg`. This model had the lowest AIC score of $-744.09$.

## Backward Stepwise Regression
```{r}
rq1_model_backward <- step(full_model, scope=list(lower=simple_model, upper=full_model), direction="backward")
```

```{r}
summary(rq1_model_backward)
```
Based on the output from the backward stepwise regression, the variables selected were `e`, `w..deg`, `Node..deg`, `q..AU`, and `P..yr`. This model had the lowest AIC score of $-746.01$.

## Lasso Regression
```{r}
# load necessary library
library(glmnet)

X <- model.matrix(MOID..AU. ~., data = sub.orbit)[, -1] # Remove the intercept column
y <- sub.orbit$MOID..AU.

# Fit the Lasso regression model

# Cross-validation to find the best lambda
cv_model <- cv.glmnet(X, y, alpha = 1)

# Get the best lambda
best_lambda <- cv_model$lambda.min
print(best_lambda)

# Fit the model with the best lambda
best_lasso_model <- glmnet(X, y, alpha = 1, lambda = best_lambda)
best_lasso_model

# Get the coefficients of the best model
coefficients <- coef(best_lasso_model)
# Convert the sparse matrix to a regular matrix
coefficients <- as.matrix(coefficients)
print(coefficients)

# Plot the coefficients against the log(lambda)
base_lasso_model <- glmnet(X, y, alpha = 1)
plot(base_lasso_model, xvar = "lambda", label = TRUE)
```

```{r}
# Define a small threshold value
threshold <- 1e-6

# Identify coefficients that are not close to zero
selected_variables <- rownames(coefficients)[abs(coefficients) > threshold]

# Print the selected variables
selected_variables
```
As seen from the output above, with a threshold of $1e-6$, the variables selected are `e`, `i..deg`, `w..deg`, `Node..deg`, `q..AU`, and `P..yr`. The rest of the cofficients decreased to 0 or close to it at the optimal lambda value of $0.001753752$.

```{r}
# define linear model with lasso chosen vars
rq1_model_lasso_chosen <- lm(MOID..AU. ~ e + i..deg. + w..deg. + Node..deg. + q..AU. + P..yr., data=sub.orbit)
summary(rq1_model_lasso_chosen)
```
# Model Assumptions
## Visualizations
### rq1_model_forward
```{r}
par(mfrow=c(2,2))
qqnorm(residuals(rq1_model_forward))
qqline(residuals(rq1_model_forward))
hist(residuals(rq1_model_forward), main="Histogram of residuals",
xlab="Residuals")
plot(residuals(rq1_model_forward), xlab="Order", ylab="Residuals")
abline(0, 0, lty=1, col="red")
plot(fitted(rq1_model_forward), residuals(rq1_model_forward), xlab="Fitted values",
ylab="Residuals")
abline(0, 0, lty=1, col="red")
```

### rq1_model_backward
```{r}
par(mfrow=c(2,2))
qqnorm(residuals(rq1_model_backward))
qqline(residuals(rq1_model_backward))
hist(residuals(rq1_model_backward), main="Histogram of residuals",
xlab="Residuals")
plot(residuals(rq1_model_backward), xlab="Order", ylab="Residuals")
abline(0, 0, lty=1, col="red")
plot(fitted(rq1_model_backward), residuals(rq1_model_backward), xlab="Fitted values",
ylab="Residuals")
abline(0, 0, lty=1, col="red")
```

### rq1_model_lasso_chosen
```{r}
par(mfrow=c(2,2))
qqnorm(residuals(rq1_model_lasso_chosen))
qqline(residuals(rq1_model_lasso_chosen))
hist(residuals(rq1_model_lasso_chosen), main="Histogram of residuals",
xlab="Residuals")
plot(residuals(rq1_model_lasso_chosen), xlab="Order", ylab="Residuals")
abline(0, 0, lty=1, col="red")
plot(fitted(rq1_model_lasso_chosen), residuals(rq1_model_lasso_chosen), xlab="Fitted values",
ylab="Residuals")
abline(0, 0, lty=1, col="red")
```












