---
title: "ISyE 6414 Final Projecct: Predicting for Near Earth Comets"
author: "Troy Allen, Roy Gabriel, Nattakorn Kittisut, Corey Smith & Moe Kyaw Thu"
date: "2024-06-13"
output: html_document
---

Let's load in the data and some necessary packages first.
```{r}
library(tidyverse)
library(caret)
library(glmnet)
library(stargazer)
comet_data=read.csv("Near-Earth_Comets_-_Orbital_Elements_20240605.csv")
comet_data
```

Checking for missing values:
```{r}
# Find nan values
missing_values <- comet_data %>% 
  summarise(across(everything(), ~ sum(is.na(.)))) %>% 
  gather(key = "Column", value = "MissingValues")

# Print nan values
#missing_values
```

We can see the missing values are as follows:

A1 = 122 missing values
A2 = 119 missing values
A3 = 150 missing values
DT = 156 missing values

Obviously can't drop all of these rows due to the sheer number of missing values we have.

```{r}
# Print out the 4 problematic columns
print(comet_data %>% select(A1..AU.d.2., A2..AU.d.2., A3..AU.d.2., DT..d.))
```
Dropping the columns A3 and DT seems like the best option for now. Currently dropping A1/A2 as well since I wanted to see what Corey has to say about the importance of these columns in our models.

```{r}
# Drop the 'A3' and 'DT' columns
comet_data_no_A3DT <- comet_data %>%
  select(-c(A1..AU.d.2., A2..AU.d.2.,A3..AU.d.2., DT..d.))

# Print the dataset without A3/DT
comet_data_no_A3DT
```

Let's check for outliers now.

```{r}
# Filter out non-numeric columns
numeric_cols <- select_if(comet_data_no_A3DT, is.numeric)

# Calculate Z-scores for each numeric column
z_scores <- numeric_cols %>% 
  mutate(across(everything(), scale))

# Define threshold
threshold <- 2

# Identify outliers based on Z-score
outliers <- comet_data_no_A3DT %>% 
  filter(rowSums(abs(z_scores) > threshold, na.rm = TRUE) > 0)

print(outliers)
```

```{r}
# Filter the outlier columns based on threshold
outlier_columns <- apply(abs(z_scores) > threshold, 1, function(x) colnames(comet_data_no_A3DT)[x])

# Print which columns the outlier lies in and the entire row associated with outlier
for(i in 1:nrow(outliers)) {
  if(length(outlier_columns[[i]]) > 0) {
    cat("Row:", i, "\n")
    cat("Outlier columns:", paste(outlier_columns[i], collapse = ", "), "\n")
    print(outliers[i, ])
  }
}

```
Wanted to compare each to the mean and stddev of each column.

```{r}
# Filter out numeric columns
numeric_data <- comet_data_no_A3DT[, sapply(comet_data_no_A3DT, is.numeric)]

# Calculate mean of each column
column_means <- colMeans(numeric_data, na.rm = TRUE)

# Calculate standard deviation of each column
column_stddev <- sapply(numeric_data, sd, na.rm = TRUE)

# Print mean and standard deviation of each column
for (i in seq_along(column_means)) {
  cat("Column:", names(column_means)[i], "\n")
  cat("Mean:", column_means[i], "\n")
  cat("Standard Deviation:", column_stddev[i], "\n\n")
}
```

Creating a standardized version of the data in case it is needed.

```{r}
# Get numeric columns from cleaned data
numeric_comet_data <- comet_data_no_A3DT[, sapply(comet_data_no_A3DT, is.numeric)]

# Get mean + sd
means <- colMeans(numeric_comet_data)
sds <- apply(numeric_comet_data, 2, sd)

# Standardize each column and print
standardized_comet_data <- scale(numeric_comet_data, center = means, scale = sds)
```

Some simple plots based on Eccentricity, perihelion distance, and orbital elements:

```{r}
ggplot(comet_data_no_A3DT, aes(x = e, y = i..deg.)) +
  geom_point() +
  labs(x = "Eccentricity", y = "Inclination (deg)", title = "Scatter Plot of Eccentricity vs. Inclination")
```

```{r}
ggplot(comet_data_no_A3DT, aes(x = e, y = q..AU.)) +
  geom_point() +
  labs(x = "Eccentricity", y = "Perihelion Distance (q AU)", title = "Scatter Plot of Eccentricity vs. Perihelion Distance") 
```

```{r}
# Filter orbital elements 
orbital_elements <- comet_data[, c("q..AU.", "Epoch..TDB.", "TP..TDB.", "e", "i..deg.", "w..deg.", "Node..deg.")]

# Create a scatterplot matrix showing relationship between q and other orbital elements
pairs(orbital_elements, pch = 21, bg = "skyblue")

#Write CSV
#write.csv(comet_data_no_A3DT, "comet_data_cleaned.csv", row.names = FALSE)
```

```{r}
#Reading data
orbit <- read.csv("comet_data_cleaned.csv")
print(head(orbit))
```

```{r}
#Loading appropriate packages for statistical analysis
library(readr)
library(ggplot2)
library(dplyr)
library(car)
library(stats)
library(lmtest)
library(gghalfnorm)
library(faraway)
library(MASS)
library(caret)
library(tidyverse)
library(caret)
library(glmnet)
```

```{r}
#RQ1: Which orbital elements best predict the minimum distance between the comet and Earth (MOID)? Are all input parameters significant? Is the MOID impacted more by the orbit eccentricity (e) or its critical angles (i, w, Node)?

#Sub-setting data for clarity in linear regression modeling
#Deleted columns include Object, Ref, and Object_name since they will only lead to NANs if included in the model.
sub.orbit <- orbit[,2:11]
rq1.lm <- lm(MOID..AU. ~., data = sub.orbit)
summary(rq1.lm)
#Interpretation: Variables namely q..AU., Node..deg., w..deg., and e are significant predictor variables
```


```{r}
#RQ1 Modal Diagnostics:
#Checking for normality
print(shapiro.test(resid(rq1.lm)))
#Interpretation: There is strong evidence that the residuals of the regression model do not come from a normally distributed population.

#Checking for multicollinearity
print(vif(rq1.lm))
#There are mainly four multicollinearity: Epoch..TDB., TP..TDB., Q..AU., and P..yr.

#Checking for Homoscedasticity/heteroskedasticity
print(ncvTest(rq1.lm))
#With p-value less than 0.05, there is heteroskedasticity in the model. 

#Checking for outlier values
cook.rq.1 <- cooks.distance(rq1.lm)
halfnorm(cook.rq.1,5,ylab="Cook’s distances")
#Interpretation: The outliers are observations 5, 158, 6, 22, and 160.
```

```{r}
# Results from tests
shapiro_w <- 0.89569
shapiro_p <- 3.234e-09

vif_values <- c(Epoch_TDB = 575.732499, TP_TDB = 575.160715, e = 11.270028, i_deg = 2.146876, 
                w_deg = 1.137160, Node_deg = 1.146094, q_AU = 5.176505, Q_AU = 152.786828, P_yr = 108.347699)

ncv_chisquare <- 21.54705
ncv_df <- 1
ncv_p <- 3.4525e-06
cook.outlie <- c(5, 158, 6, 22, 160)

# Creating the table
result_table <- data.frame(
  Test = c(
    "Shapiro-Wilk normality test", 
    rep("Variance Inflation Factor (VIF)", length(vif_values)), 
    "Non-constant Variance Score Test", 
    "Cook's Distance"
  ),
  Result = c(
    paste("W:", shapiro_w, ", p-value:", format(shapiro_p, scientific = TRUE)),
    paste(names(vif_values), ":", round(vif_values, 6)),
    paste("Chisquare:", ncv_chisquare, ", Df:", ncv_df, ", p:", format(ncv_p, scientific = TRUE)),
    paste("Outliers:", paste(cook.outlie, collapse = ", "))
  )
)

# Printing the table
stargazer(result_table, type = "text", summary = FALSE, rownames = FALSE, out = "stargazer.txt")


```

```{r}
#RQ1: Further modeling: Stepwise Regression
simple_model <- lm(MOID..AU. ~ 1, data = sub.orbit)
full_model <- lm(MOID..AU. ~., data = sub.orbit)
```

```{r}
#RQ1 Stepwise Regression
#Forward Stepwise Regression
rq1_model_forward <- step(simple_model, scope=list(lower=simple_model, upper=full_model), direction="forward")
summary(rq1_model_forward)
#Interpretation: Based on the output of the forward stepwise regression, the variables selected were `Q..AU`, `P..yr`, `q..AU`, `e`, `Node..deg`, and `w..deg`. This model had the lowest AIC score of -744.09.

#Visualization for Forward Stepwise Regression
par(mfrow=c(2,2))
qqnorm(residuals(rq1_model_forward))
qqline(residuals(rq1_model_forward))
hist(residuals(rq1_model_forward), main="Histogram of residuals",
xlab="Residuals")
plot(residuals(rq1_model_forward), xlab="Order", ylab="Residuals")
abline(0, 0, lty=1, col="red")
plot(fitted(rq1_model_forward), residuals(rq1_model_forward), xlab="Fitted values",
ylab="Residuals")
abline(0, 0, lty=1, col="red")
```

```{r}
#Backward Stepwise Regression
rq1_model_backward <- step(full_model, scope=list(lower=simple_model, upper=full_model), direction="backward")
summary(rq1_model_backward)
#Interpretation: Based on the output from the backward stepwise regression, the variables selected were `e`, `w..deg`, `Node..deg`, `q..AU`, and `P..yr`. This model had the lowest AIC score of -746.01.

#Visualization for Backward Stepwise Regression
par(mfrow=c(2,2))
qqnorm(residuals(rq1_model_backward))
qqline(residuals(rq1_model_backward))
hist(residuals(rq1_model_backward), main="Histogram of residuals",
xlab="Residuals")
plot(residuals(rq1_model_backward), xlab="Order", ylab="Residuals")
abline(0, 0, lty=1, col="red")
plot(fitted(rq1_model_backward), residuals(rq1_model_backward), xlab="Fitted values",
ylab="Residuals")
abline(0, 0, lty=1, col="red")
```

```{r}
#Lasso Regression for RQ1
X <- model.matrix(MOID..AU. ~., data = sub.orbit)[, -1] # Remove the intercept column
y <- sub.orbit$MOID..AU.

# Fit the Lasso regression model
# Cross-validation to find the best lambda
cv_model <- cv.glmnet(X, y, alpha = 1)

# Get the best lambda
best_lambda <- cv_model$lambda.min
print(best_lambda)

# Fit the model with the best lambda
best_lasso_model <- glmnet(X, y, alpha = 1, lambda = best_lambda)
best_lasso_model

# Get the coefficients of the best model
coefficients <- coef(best_lasso_model)
# Convert the sparse matrix to a regular matrix
coefficients <- as.matrix(coefficients)
print(coefficients)

# Plot the coefficients against the log(lambda)
base_lasso_model <- glmnet(X, y, alpha = 1)
plot(base_lasso_model, xvar = "lambda", label = TRUE)
```

```{r}
# Define a small threshold value
threshold <- 1e-6

# Identify coefficients that are not close to zero
selected_variables <- rownames(coefficients)[abs(coefficients) > threshold]

# Print the selected variables
selected_variables

#Interpretation: As seen from the output above, with a threshold of $1e-6$, the variables selected are `e`, `i..deg`, `w..deg`, `Node..deg`, `q..AU`, and `P..yr`. The rest of the cofficients decreased to 0 or close to it at the optimal lambda value of $0.001753752$.
```

```{r}
#linear Regression Modeling with lasso chosen vars
rq1_model_lasso_chosen <- lm(MOID..AU. ~ e + i..deg. + w..deg. + Node..deg. + q..AU. + P..yr., data=sub.orbit)
summary(rq1_model_lasso_chosen)
#Interpretation: Other than i..deg., all the variables are significant predictors for MOID..AU..

#Visualization for regression modeling with chosen Lasso variables
par(mfrow=c(2,2))
qqnorm(residuals(rq1_model_lasso_chosen))
qqline(residuals(rq1_model_lasso_chosen))
hist(residuals(rq1_model_lasso_chosen), main="Histogram of residuals",
xlab="Residuals")
plot(residuals(rq1_model_lasso_chosen), xlab="Order", ylab="Residuals")
abline(0, 0, lty=1, col="red")
plot(fitted(rq1_model_lasso_chosen), residuals(rq1_model_lasso_chosen), xlab="Fitted values",
ylab="Residuals")
abline(0, 0, lty=1, col="red")
```

```{r}
#RQ2: Do the minimum (q) and maximum (Q) distances between the comet and the Sun have an impact on the orbital period (P)? Additionally, are these minimum and maximum distance values correlated or do they depend on some or all other orbital elements?

rq2.lm <- lm(P..yr. ~ q..AU. + Q..AU. + I(q..AU. * Q..AU.), data = sub.orbit)
summary(rq2.lm)
#Interpretation: The maximum distance (Q) and the interaction between minimum (q) and maximum (Q) distances significantly impact the orbital period of the comet, while the minimum distance (q) alone does not have a significant orbital period (P).

#Further correlation test
cor(sub.orbit$q..AU., sub.orbit$Q..AU.)
#Interpretation: With the correlation value of 0.044, there is a weak positive correlation between the two.

#More correlation testing procedures: Overall correlation
cor(sub.orbit)
#Interpretation: Based on the testing procedures, it can be seen that e has the highest correlation with both q and Q.
#This means that these two are also dependent on q and Q.

#Based on this further regression modeling is conducted:
rq2.lm.2 <- lm(P..yr. ~ q..AU. + Q..AU. + I(q..AU. * Q..AU.) + e, data = sub.orbit)
summary(rq2.lm.2)
#Interpretation: all of the variables are now significant.
```

```{r}
#RQ2 Model Diagnostics:
#Checking for normality
print(shapiro.test(resid(rq2.lm)))
print(shapiro.test(resid(rq2.lm.2)))
#Both models are not normally distributed.

#Checking for multicollinearity
print(vif(rq2.lm))
print(vif(rq2.lm.2))
#In both mode, Q and the interaction terms between q and Q have high VIFs.

#Checking for Homoscedasticity/heteroskedasticity
print(ncvTest(rq2.lm))
print(ncvTest(rq2.lm.2))
#With p-value less than 0.05, there is heteroskedasticity in both of the model. 

#Checking for leverage points
cook.rq.2.lim <- cooks.distance(rq2.lm)
cook.rq.2.lm.2 <- cooks.distance(rq2.lm.2)
halfnorm(cook.rq.2.lim,5,ylab="Cook’s distances")
halfnorm(cook.rq.2.lm.2,ylab = "Cook's distances")
#The two outliers are 159 and 123.
```

```{r}
#RQ3.Is there any relationship between the two response variables, MOID and P? Is there a way to predict the likelihood and/or frequency of a Near-Earth comet sighting? Provide justifications using the results of the regression analysis.
rq3.lm <- lm(MOID..AU. ~ P..yr., data = sub.orbit)
summary(rq3.lm)
#Interpretation: There is a really high significant correlation between MOID and P although the coefficient is very low at 0.0012.

#Checking for correlation
cor(sub.orbit$MOID..AU., sub.orbit$P..yr.)
#Interpretation: There is a decently high correlation between the two as well.

#Further correlation procedures
#Based on the correlation modeling above from RQ2, it can be seen that MOID is positively correlated with P (0.34) and Q (0.29).Meanwhile, P is positively and strongly correlation with Q (0.98) and e (0.60).

#Further regression modeling using Q, and e as additional variables
rq3.lm.2 <- lm(MOID..AU. ~ P..yr. + Q..AU. + e, data = sub.orbit)
summary(rq3.lm.2)
#Interpretation: Results make sense since MOID is strongly correlated with P and Q, not with e.

#Predicting for likelihood using logistic regression
#Creating binary for Near Earth comets
sub.orbit$Near <- ifelse(sub.orbit$MOID..AU. < 0.05, 1, 0) #Values which are less than 0.05
rq3.lm.3 <- glm(Near ~ P..yr. + q..AU. + Q..AU. + e + i..deg. + w..deg. + Node..deg., family = binomial, data = sub.orbit)
summary(rq3.lm.3)
#Interpretation: Only q, e, and Node degrees are significant predictors
```

```{r}
#RQ3 Model Diagnostics:
#Checking for normality
print(shapiro.test(resid(rq3.lm)))
print(shapiro.test(resid(rq3.lm.2)))
print(shapiro.test(resid(rq3.lm.3)))
#Interpretation: None of the models' residuals are normally distributed.

#Checking for multicollinearity; the first model for rq3 is not checked since there's only one predictor variable
print(vif(rq3.lm.2))
print(vif(rq3.lm.3))
#In both models, P and Q have high VIF values thus big multicollinearity.

#Checking for Homoscedasticity/heteroskedasticity
print(ncvTest(rq3.lm))
print(ncvTest(rq3.lm.2))
#Interpretation: There is heteroskedasticity in both models

#Checking for Homoscedasticity/heteroskedasticity for logisitic regression model
print(bptest(rq3.lm.3))
#Interpretation: There is heteroskedasticity in the logistic regression model as well.

#Checking for outlier values
cook.rq.3 <- cooks.distance(rq3.lm)
cook.rq.3.2 <- cooks.distance(rq3.lm.2)
cook.rq.3.3 <- cooks.distance(rq3.lm.3)
halfnorm(cook.rq.3,5,ylab="Cook’s distances")
halfnorm(cook.rq.3.2,5,ylab="Cook’s distances")
halfnorm(cook.rq.3.3,5,ylab="Cook’s distances")
#In the original model, the outliers are 123, 98, and 159.
#In the second model, the outliers are 159, 158, 98, and 123.
#In the logistic model, the outliers are 123, 98, and 22.
#In all three models, it seems that 123 and 98 are outliers.
```










